---
title: "Getting started with rwevalidate"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with rwevalidate}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

`rwevalidate` validates an **already-instantiated** patient cohort on an OMOP
CDM database for real-world-evidence (RWE) use, and produces a structured HTML +
JSON report aligned with FDA RWE guidance (Relevance and Reliability) and the
HARPER protocol framework. It does **not** generate cohorts -- the cohort
definition (SQL or ATLAS) stays upstream.

The code below is not executed when the vignette is built (it needs a live
database); it is the exact sequence you would run against your own CDM.

## 1. Connect

`cdm_connect()` opens a PostgreSQL connection and verifies the required clinical
tables exist in `cdm_schema` before returning.

```{r}
library(rwevalidate)

con <- cdm_connect(
  host       = "localhost",
  port       = 5432,
  dbname     = "your_db",
  user       = "your_user",
  password   = "your_password",
  cdm_schema = "mimic_cdm"
)
```

Many CDM builds are **split-schema**: clinical tables in one schema and the
vocabulary (`concept`, `concept_ancestor`) in another. Pass the vocabulary
schema via `vocab_schema` (default `"vocab"`).

## 2. Instantiate a cohort (upstream)

`rwevalidate` reads a cohort table with the columns `subject_id`,
`cohort_definition_id`, `cohort_start_date`, `cohort_end_date`. For example, a
heart-failure cohort using SNOMED concept 316139 and its descendants:

```{r}
DBI::dbExecute(con, "
  CREATE TABLE results.my_cohort AS
  SELECT person_id AS subject_id, 1 AS cohort_definition_id,
         MIN(condition_start_date) AS cohort_start_date,
         MAX(condition_end_date)   AS cohort_end_date
  FROM mimic_cdm.condition_occurrence
  WHERE condition_concept_id IN (
    SELECT descendant_concept_id FROM vocab.concept_ancestor
    WHERE ancestor_concept_id = 316139)
  GROUP BY person_id")
```

## 3. Validate in one call

`validate_cohort()` runs the modules and writes the report. Supply an open `con`
or the connection arguments directly.

```{r}
results <- validate_cohort(
  cdm_schema    = "mimic_cdm",
  cohort_table  = "results.my_cohort",
  cohort_id     = 1,
  concept_ids   = 316139,   # enables Module 1 (concept coverage)
  comparator_id = 2,        # enables Module 4 (covariate feasibility)
  vocab_schema  = "vocab",
  output_dir    = "./validation_report",
  con           = con
)
```

This writes `validation_report.html` and `validation_results.json` to
`output_dir`, and returns a list with `results` (per-module output),
`report` (paths + traffic-light summary), and `flags`.

```{r}
results$report$check_summary   # one row per check: pass / warn / fail
results$flags                  # all WARN:/FAIL: messages
```

## 4. Run modules individually

Each module is also callable on its own when you only need one view.

```{r}
run_attrition(con, "mimic_cdm", "results.my_cohort", cohort_id = 1,
              vocab_schema = "vocab")

run_density(con, "mimic_cdm", "results.my_cohort", cohort_id = 1)

run_concepts(con, "mimic_cdm", "results.my_cohort", cohort_id = 1,
             concept_ids = 316139, vocab_schema = "vocab")

run_covariates(con, "mimic_cdm", "results.my_cohort", cohort_id = 1,
               comparator_id = 2, vocab_schema = "vocab")
```

## 5. Interpreting the traffic light

Every check returns `"pass"` (green), `"warn"` (amber), or `"fail"` (red).
Thresholds are documented arguments -- for example prior-observation coverage
warns below 70% and fails below 50%, and covariates with `|SMD| > 0.1` are
flagged as imbalanced. Tune them to your study's requirements.

```{r}
cdm_disconnect(con)
```
